---
title: "Chapter 04 - 3. Gaussian model of height"
format: html
editor: visual
---

## Gaussian model of height

## The data

```{r}
#| message: false
library(rethinking)
data(Howell1)
d <- Howell1
d2 <- d[ d$age >= 18 , ]
```

```{r}
str( d )
```

```{r}
precis( d )
```

## The model

$$
\begin{align}
&h_i \sim \text{Normal}(\mu, \sigma) \qquad &\text{[likelihood]} \\
&\mu \sim \text{Normal}(178, 20)  \qquad &\text{[ $\mu$ prior]} \\
&\sigma \sim \text{Uniform}(0, 50)  \qquad &\text{[$\sigma$ prior]} 
\end{align}
$$

### Prior distributions

```{r}
curve( dnorm( x , 178 , 20 ) , from=100 , to=250 )
```

```{r}
curve( dunif( x , 0 , 50 ) , from=-10 , to=60 )
```

### Prior Predictive simulation

```{r}
sample_mu <- rnorm( 1e4 , 178 , 20 )
sample_sigma <- runif( 1e4 , 0 , 50 )
prior_h <- rnorm( 1e4 , sample_mu , sample_sigma )

dens( prior_h )
```

## **Grid approximation of the posterior distribution**

```{r}
mu.list <- seq( from=150, to=160 , length.out=100 )
sigma.list <- seq( from=7 , to=9 , length.out=100 )
post <- expand.grid( mu=mu.list , sigma=sigma.list )
post$LL <- sapply(
  1:nrow(post),
  function(i) sum(
    dnorm(d2$height, post$mu[i], post$sigma[i], log=TRUE)
    )
  )
post$prod <- post$LL + dnorm( post$mu , 178 , 20 , TRUE ) + dunif( post$sigma , 0 , 50 , TRUE )

post$prob <- exp( post$prod - max(post$prod) ) # This is not scaled (They are probability weights)
```

```{r}
contour_xyz( post$mu , post$sigma , post$prob )
```

```{r}
image_xyz( post$mu , post$sigma , post$prob )
```

### Sampling from the posterior

```{r}
sample.rows <- sample(
  1:nrow(post), size=1e4, replace=TRUE, prob=post$prob
  )
sample.mu <- post$mu[ sample.rows ]
sample.sigma <- post$sigma[ sample.rows ]
```

```{r}
plot(sample.mu , sample.sigma , cex=0.5 , pch=16 , col=col.alpha(rangi2,0.1) )
```

```{r}
dens( sample.mu )
```

```{r}

dens( sample.sigma )
```

```{r}
PI( sample.mu )
PI( sample.sigma )
```

### **Overthinking: Sample size and the normality of *σ*'s posterior**.

```{r}
d3 <- sample( d2$height , size=20 )

mu.list <- seq( from=150, to=170 , length.out=200 )
sigma.list <- seq( from=4 , to=20 , length.out=200 )
post2 <- expand.grid( mu=mu.list , sigma=sigma.list )
post2$LL <- sapply( 1:nrow(post2) , function(i) sum( dnorm( d3 , mean=post2$mu[i] , sd=post2$sigma[i] ,log=TRUE ) ) )
post2$prod <- post2$LL + dnorm( post2$mu , 178 , 20 , TRUE ) +
dunif( post2$sigma , 0 , 50 , TRUE )
post2$prob <- exp( post2$prod - max(post2$prod) )
sample2.rows <- sample( 1:nrow(post2) , size=1e4 , replace=TRUE, prob=post2$prob )
sample2.mu <- post2$mu[ sample2.rows ]
sample2.sigma <- post2$sigma[ sample2.rows ]
plot( sample2.mu , sample2.sigma , cex=0.5 , col=col.alpha(rangi2,0.1) , xlab="mu" , ylab="sigma" , pch=16 )
```

```{r}
dens( sample2.sigma , norm.comp=TRUE )
```

## **Finding the posterior distribution with `quap`**

> Now we leave grid approximation behind and move on to one of the great engines of applied statistics, the **QUADRATIC APPROXIMATION**. Our interest in quadratic approximation, recall, is as a handy way to quickly make inferences about the shape of the posterior. The posterior's peak will lie at the **MAXIMUM A POSTERIORI** estimate (MAP), and we can get a useful image of the posterior's shape by using the quadratic approximation of the posterior distribution at this peak.

```{r}
library(rethinking)
data(Howell1)
d <- Howell1
d2 <- d[ d$age >= 18 , ]
```

```{r}
flist <- alist(
  height ~ dnorm( mu , sigma ),
  mu ~ dnorm( 178 , 20 ),
  sigma ~ dunif( 0 , 50 )
)
```

```{r fit_model}
m4.1 <- quap( flist , data=d2 )
```

```{r}
precis( m4.1 )
```

### **Overthinking: Start values for `quap`**. `quap`

```{r}
start <- list(
  mu=mean(d2$height),
  sigma=sd(d2$height)
)
m4.1a <- quap( flist , data=d2 , start=start )
precis( m4.1a )
```

### **Sampling from a `quap`.**

```{r}
post <- extract.samples( m4.1 , n=1e4 )
head(post)
```

```{r}
precis(post)
```
