---
title: "Homework - Week 07"
format:
  html:
    embed-resources: true
---

```{r}
#| warning: false
#| message: false
library(rethinking)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(dagitty)
library(patchwork)
```


## Problem 01

The data in `data(bangladesh)` are 1934 women from the 1989 Bangladesh Fertility Survey. For each woman, we know which `district` she lived in (`D`), her number
of living.children (`K`), her age.centered (`A`), whether she lived in an urban center (`U`),
and finally whether or not she used contraception (use.contraception) (`C`).
In the lecture, we estimated the total effect of urban residence on contraceptive
use. Using the DAG from lecture, or your own version if you prefer, **now estimate only the direct effect of urban residence on contraceptive use**.

```{r}
data(bangladesh)
d <- bangladesh %>% 
  as_tibble %>% 
  transmute(
    C = use.contraception,
    D = as.integer(district),
    U = as.integer(urban),
    A = age.centered,
    K = as.numeric(living.children) # The min number of living children in the data set is 1
  )
d
```


```{r}
exdag <- dagitty( "dag {
    A -> C
    K -> C
    U -> C
    D -> C
    A -> K
    U -> K
    D -> U
}")
coordinates( exdag ) <- list( 
  x=c(C=0,A=-1,D=1,K=-0.5,U=0.5) , 
  y=c(C=0,A=0.2,D=0.2,K=0.5,U=0.5) )
drawdag(exdag)
```

Total effect adjust set
```{r}
dagitty::adjustmentSets(exdag, exposure = "U", outcome = "C", effect = "total")
```
Statistica model

$$
\begin{align}
C_i &\sim \text{Bernoulli}(p_i) \\
\text{logit}(p_i) &= \alpha_{D[i]} + \beta_{D[i]}U_i \\
\alpha &= \bar{\alpha} + v_{\_,1} \\
\beta &= \bar{\beta} + v_{\_,2} \\
v &= (\text{diag}(\sigma)\mathbf{L}\mathbf{Z})^T \\
\mathbf{Z}_{j,k} &\sim \text{Normal}(0,1) \\
\mathbf{\sigma} &\sim \text{Exponential}(1) \\
\mathbf{L} &\sim \text{LKJCorrCholesky}(4) \\
\end{align}
$$

```{r}
#| output: false
#| message: false
#| cache: true
m1_total <- ulam(
  alist(
    C ~ bernoulli(p),
    logit(p) <- a[D] + b[D]*U,
    # define effects using other parameters
    # this is the non-centered Cholesky machine
    transpars> vector[61]:a <<- abar[1] + v[,1],
    transpars> vector[61]:b <<- abar[2] + v[,2],
    transpars> matrix[61,2]:v <- compose_noncentered( sigma , L_Rho , Z ),
    # priors - note that none have parameters inside them
    # that is what makes them non-centered
    matrix[2,61]:Z ~ normal( 0 , 1 ),
    vector[2]:abar ~ normal(0,1),
    cholesky_factor_corr[2]:L_Rho ~ lkj_corr_cholesky( 4 ),
    vector[2]:sigma ~ exponential(1),
    # convert Cholesky to Corr matrix
    gq> matrix[2,2]:Rho <<- Chol_to_Corr(L_Rho)
  ) , data=d , chains=4 , cores=4 )
```

Check the value for U
```{r}
precis(m1_total, depth = 2, pars = "abar")
```

```{r}
total_precis <- precis(m1_total, depth = 2)
as_tibble(total_precis) %>% 
  mutate(param = row.names(total_precis)) %>% 
  select(param, n_eff, Rhat4) %>% 
  arrange(n_eff)
```

### What is the total effect of urban residence on contraceptive use?

```{r}
total_posterior <- extract.samples(m1_total)
p_total_samples <- inv_logit(total_posterior$a + total_posterior$b) * 100

mean_p <- p_total_samples %>% apply(MARGIN=2, FUN=mean)
p89interval <- p_total_samples %>% apply(MARGIN=2, FUN=PI) %>% t
total_to_plot <- tibble(
  p = mean_p,
  p5 = p89interval[,1],
  p94 = p89interval[,2]
) %>% 
  mutate(D = row_number())

total_effect_plot <- ggplot(total_to_plot) +
  geom_segment(
    aes(x=D, xend=D, y=p5, yend=p94), 
    alpha = 0.4, linewidth=2, lineend = "round"
  ) +
  geom_point(aes(x=D, y=p)) +
  geom_hline(aes(yintercept=mean(p_total_samples)), linetype=2) +
  scale_y_continuous(limits = c(0, 100))+
  labs(x="District", y="Percentual points (pp)", 
       title = "Total effect of Urban residence on contraceptive us per district") +
  theme_minimal()
total_effect_plot
```

```{r}
total_posterior <- extract.samples(m1_total)
p_samples_urban <- inv_logit(total_posterior$a + total_posterior$b) * 100

mean_p <- p_samples_urban %>% apply(MARGIN=2, FUN=mean)
p89interval <- p_samples_urban %>% apply(MARGIN=2, FUN=PI) %>% t
total_urban_to_plot <- tibble(
  p = mean_p,
  p5 = p89interval[,1],
  p94 = p89interval[,2]
) %>% 
  mutate(D = row_number(),
         U= "urban")


p_samples_rural <- inv_logit(total_posterior$a) * 100

mean_p <- p_samples_rural %>% apply(MARGIN=2, FUN=mean)
p89interval <- p_samples_rural %>% apply(MARGIN=2, FUN=PI) %>% t
total_rural_to_plot <- tibble(
  p = mean_p,
  p5 = p89interval[,1],
  p94 = p89interval[,2]
) %>% 
  mutate(D = row_number(),
         U = "rural")

total_effect_by_category_plot <- ggplot(bind_rows(total_rural_to_plot, total_urban_to_plot)) +
  geom_segment(
    aes(x=D, xend=D, y=p5, yend=p94, color=U), 
    alpha = 0.4, linewidth=2, lineend = "round"
  ) +
  geom_point(aes(x=D, y=p, color=U)) +
  geom_hline(aes(yintercept=mean(c(p_samples_rural, p_samples_urban))), linetype=2) +
  scale_y_continuous(limits = c(0, 100)) +
  facet_grid( U ~ .) +
  labs(x="District", y="Percentual points (pp)", 
       title = "Total effect of Urban residence on contraceptive us per district") +
  theme_minimal()
total_effect_by_category_plot
```


### Direct effect model


Direct effect adjustment set
```{r}
dagitty::adjustmentSets(exdag, exposure = "U", outcome = "C", effect = "direct")
```

Model
$$
\begin{align}
C_i &\sim \text{Bernoulli}(p_i) \\
\text{logit}(p_i) &= \alpha_{D[i]} + \beta^U_{D[i]}U_i + \beta^A_{D[i]}A_i + \beta^K\sum_{j=0}^{K_i-1}\delta_j \\
\alpha &= \bar{\alpha} + v_{\_,1} \\
\beta^U &= \bar{\beta}^U + v_{\_,2} \\
\beta^A &= \bar{\beta}^A + v_{\_,3} \\
\beta^K &= \bar{\beta}^K + v_{\_,4} \\
v &= (\text{diag}(\mathbf{\sigma})\mathbf{L}\mathbf{Z})^T \\
\mathbf{Z}_{j,k} &\sim \text{Normal}(0,1) \\
\mathbf{\sigma} &\sim \text{Exponential}(1) \\
\mathbf{L} &\sim \text{LKJCorrCholesky}(4) \\
\mathbf{\delta} &\sim \text{Dirichlet}(a)
\end{align}
$$
*Notes*
- The effect of $A$ is expected to vary by district
- $\beta^K$ the "maximum effect" of having children is expected to vary by district
- $\delta$ is the proportion of the maximum effect and their values aren't assumed to be associated with the district.

#### Model with age and kids

```{r direct_complete_model}
#| output: false
#| message: false
#| cache: true

d_list <- list(
  C = d$C,
  D = d$D,
  U = d$U,
  A = d$A,
  K = d$K
)
d_list$prior_dirichlet <- rep(2, 3)

m1_direct_complete <- ulam(
  alist(
    C ~ bernoulli(p),
    logit(p) <- a[D] + 
      bU[D]*U + 
      bA[D]*A + 
      bK*sum(delta_j[1:K]),
    # define effects using other parameters
    # this is the non-centered Cholesky machine
    transpars> vector[61]:a <<- abar[1] + v[,1],
    transpars> vector[61]:bU <<- abar[2] + v[,2],
    transpars> vector[61]:bA <<- abar[3] + v[,3],
    transpars> matrix[61,3]:v <- compose_noncentered( sigma , L_Rho , Z ),
    # priors - note that none have parameters inside them
    # that is what makes them non-centered
    matrix[3,61]:Z ~ normal( 0 , 1 ),
    vector[3]:abar ~ normal(0,1),
    cholesky_factor_corr[3]:L_Rho ~ lkj_corr_cholesky( 4 ),
    vector[3]:sigma ~ exponential(1),
    vector[4]: delta_j <<- append_row(0 , delta),
    simplex[3]: delta ~ dirichlet(prior_dirichlet),
    bK ~ normal(0, 1),
    # convert Cholesky to Corr matrix
    gq> matrix[3,3]:Rho <<- Chol_to_Corr(L_Rho)
  ) , data=d_list , chains=4 , cores=4, log_lik = T)
```

```{r}
precis(m1_direct_complete, depth = 2, pars = "abar")
```

```{r}
direct_precis <- precis(m1_direct_complete, depth = 2)
to_plot <- as_tibble(direct_precis) %>% 
  mutate(param = row.names(direct_precis)) %>% 
  select(param, n_eff, Rhat4) %>% 
  arrange(n_eff)
to_plot
```

```{r}
ggplot(data=to_plot) +
  geom_point(aes(x=n_eff, y=Rhat4)) +
  geom_vline(aes(xintercept=250), color="red") +
  geom_hline(aes(yintercept=1), color="black")
```

```{r}
#| eval: false
#| include: false
traceplot(m1_direct_complete)
trankplot(m1_direct_complete)
```

```{r}
precis(m1_direct_complete, depth = 2, pars = c("bU", "bA", "bK","delta")) %>% head
```

### What is the direct effect of urban residence on contraceptive use?

```{r}
posterior <- extract.samples(m1_direct_complete)
p_direct_samples <- inv_logit(posterior$a + posterior$bU) * 100

mean_p <- p_direct_samples %>% apply(MARGIN=2, FUN=mean)
p89interval <- p_direct_samples %>% apply(MARGIN=2, FUN=PI) %>% t
direct_to_plot <- tibble(
  p = mean_p,
  p5 = p89interval[,1],
  p94 = p89interval[,2]
) %>% 
  mutate(D = row_number())
```

```{r}
direct_effect_plot <- ggplot(direct_to_plot) +
  geom_segment(
    aes(x=D, xend=D, y=p5, yend=p94), 
    alpha = 0.4, linewidth=2, lineend = "round"
  ) +
  geom_point(aes(x=D, y=p)) +
  geom_hline(aes(yintercept=mean(p_direct_samples)), linetype=2) +
  scale_y_continuous(limits = c(0, 100))+
  labs(x="District", y="Percentual points (pp)", 
       title = "Direct effect of Urban residence 
       on contraceptive us per district for Avg A and 1 child") +
  theme_minimal()
(total_effect_plot / direct_effect_plot)
```

```{r totalvsdirect}

p_contrast <- p_direct_samples - p_total_samples

mean_p <- p_contrast %>% apply(MARGIN=2, FUN=mean)
p89interval <- p_contrast %>% apply(MARGIN=2, FUN=PI) %>% t
contrast_to_plot <- tibble(
  p = mean_p,
  p5 = p89interval[,1],
  p94 = p89interval[,2]
) %>% 
  mutate(D = row_number())


ggplot(contrast_to_plot) +
  geom_segment(
    aes(x=D, xend=D, y=p5, yend=p94), 
    alpha = 0.4, linewidth=2, lineend = "round"
  ) +
  geom_point(aes(x=D, y=p)) +
  geom_hline(aes(yintercept=mean(p_contrast)), linetype=2) +
  # scale_y_continuous(limits = c(0, 100))+
  labs(x="District", y="Percentual points (pp)", 
       title = "Contrast between the direct and total effect") +
  theme_minimal()

precis(contrast_to_plot)
```


## Problem 02

Using the same DAG as before, now estimate the effect of number of surviving
children on contraceptive use. Obviously contraceptive use can also influence the
number of surviving children. But leave that issue aside for the moment (it will
return in the optional challenge further down).


```{r}
exdag <- dagitty( "dag {
    A -> C
    K -> C
    U -> C
    D -> C
    A -> K
    U -> K
    D -> U
}")
coordinates( exdag ) <- list( 
  x=c(C=0,A=-1,D=1,K=-0.5,U=0.5) , 
  y=c(C=0,A=0.2,D=0.2,K=0.5,U=0.5) )
drawdag(exdag)
```

Total effect adjust set
```{r}
dagitty::adjustmentSets(exdag, exposure = "K", outcome = "C", effect = "total")
```

```{r}
library(stringr)
m2_posterior <- extract.samples(m1_direct_complete)
one_kid_samples <- with(m2_posterior, inv_logit(a + bU)) * 100
kids <- c(1:4)
nkids <- length(kids)

bk_to_plot <- map_df(
  kids, 
  .f=function(k){
    mbK <- m2_posterior$bK %>% rep(61) %>% matrix(ncol=61)
    bK_samples <- with(
      m2_posterior,
      if(k == 1){
        inv_logit(a + bU) * 100
      }else{
        inv_logit(a + bU + mbK * apply(cbind(0,delta)[,1:k],1, sum)) * 100
      }
    )
    # Build data frame
    mean_p <- bK_samples %>% apply(MARGIN=2, FUN=mean)
    p89interval <- bK_samples %>% apply(MARGIN=2, FUN=PI) %>% t
    bk_tmp <- tibble(
      p = mean_p,
      p5 = p89interval[,1],
      p94 = p89interval[,2]
    ) %>% 
      mutate(D = row_number(), 
             xpos = str_pad((D-1) * nkids + k, 3, pad = 0, ), 
             K = as.character(k))
    return(bk_tmp)
  }
)


# Plot
bk_effect_plot <- ggplot(bk_to_plot) +
  geom_segment(
    aes(x=D, xend=D, y=p5, yend=p94, color=K), 
    alpha = 0.4, linewidth=2, lineend = "round"
  ) +
  geom_point(aes(x=D, y=p, color=K)) +
  geom_hline(aes(yintercept=mean(one_kid_samples)), linetype=2) +
  scale_y_continuous(limits = c(0, 100)) +
  facet_grid(K ~.) +
  labs(x="District", y="Percentual points (pp)", 
       title = "Direct effect of Surviving kids on contraceptive per district") +
  theme_minimal()
bk_effect_plot

```

```{r}
#| fig-height: 15
bk_effect_plot <- ggplot(bk_to_plot) +
  geom_segment(
    aes(x=xpos, xend=xpos, y=p5, yend=p94, color=K),
    alpha = 0.4, linewidth=2, lineend = "round"
  ) +
  geom_point(aes(x=xpos, y=p, color=K)) +
  geom_hline(aes(yintercept=mean(one_kid_samples)), linetype=2) +
  scale_y_continuous(limits = c(0, 100)) +
  facet_wrap(D ~., scales = "free_x") +
  labs(x="District", y="Percentual points (pp)", 
       title = "Direct effect of Surviving kids on contraceptive per district") +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(), 
        axis.text.x = element_blank())
bk_effect_plot
```

## Problem 03

Now let the causal effect of children vary by district. Incorporate this new district feature into the same multivariate prior that contains the urban/rural features. How much do districts vary in how surviving children are associated with contraceptive use?


Model
$$
\begin{align}
C_i &\sim \text{Bernoulli}(p_i) \\
\text{logit}(p_i) &= \alpha_{D[i]} + \beta^U_{D[i]}U_i + \beta^A_{D[i]}A_i + \beta^K_{D[i]}\sum_{j=0}^{K_i-1}\delta_j \\
\alpha &= \bar{\alpha} + v_{\_,1} \\
\beta^U &= \bar{\beta}^U + v_{\_,2} \\
\beta^A &= \bar{\beta}^A + v_{\_,3} \\
\beta^K &= \bar{\beta}^K + v_{\_,4} \\
v &= (\text{diag}(\mathbf{\sigma})\mathbf{L}\mathbf{Z})^T \\
\mathbf{Z}_{j,k} &\sim \text{Normal}(0,1) \\
\mathbf{\sigma} &\sim \text{Exponential}(1) \\
\mathbf{L} &\sim \text{LKJCorrCholesky}(4) \\
\mathbf{\delta} &\sim \text{Dirichlet}(a)
\end{align}
$$

```{r kids_district_model}
#| output: false
#| message: false
#| cache: true

d_list <- list(
  C = d$C,
  D = d$D,
  U = d$U,
  A = d$A,
  K = d$K
)
d_list$prior_dirichlet <- rep(2, 3)

m3_direct_complete <- ulam(
  alist(
    C ~ bernoulli(p),
    logit(p) <- a[D] + 
      bU[D]*U + 
      bA[D]*A + 
      bK[D]*sum(delta_j[1:K]),
    # define effects using other parameters
    # this is the non-centered Cholesky machine
    transpars> vector[61]:a <<- abar[1] + v[,1],
    transpars> vector[61]:bU <<- abar[2] + v[,2],
    transpars> vector[61]:bA <<- abar[3] + v[,3],
    transpars> vector[61]:bK <<- abar[4] + v[,4],
    transpars> matrix[61,4]:v <- compose_noncentered( sigma , L_Rho , Z ),
    # priors - note that none have parameters inside them
    # that is what makes them non-centered
    matrix[4,61]:Z ~ normal( 0 , 1 ),
    vector[4]:abar ~ normal(0,1),
    cholesky_factor_corr[4]:L_Rho ~ lkj_corr_cholesky( 4 ),
    vector[4]:sigma ~ exponential(1),
    vector[4]: delta_j <<- append_row(0 , delta),
    simplex[3]: delta ~ dirichlet(prior_dirichlet),
    # convert Cholesky to Corr matrix
    gq> matrix[4,4]:Rho <<- Chol_to_Corr(L_Rho)
  ) , data=d_list , chains=4 , cores=4, log_lik = T)
```

```{r}
# Model without varying children effect by district
precis(m1_direct_complete, depth = 2, pars = c("abar", "bK"))
# Model with varying children effect by district
precis(m3_direct_complete, depth = 2, pars = "abar")
```


```{r}
library(stringr)
m3_posterior <- extract.samples(m3_direct_complete)
one_kid_samples <- with(m3_posterior, inv_logit(a + bU)) * 100
kids <- c(1:4)
nkids <- length(kids)
bk_to_plot <- map_df(
  kids, 
  .f=function(k){
    bK_samples <- with(
      m3_posterior,
      if(k == 1){
        inv_logit(a + bU) * 100
      }else{
        inv_logit(a + bU + bK * apply(cbind(0,delta)[,1:k],1, sum)) * 100
      }
    )
    # Build data frame
    mean_p <- bK_samples %>% apply(MARGIN=2, FUN=mean)
    p89interval <- bK_samples %>% apply(MARGIN=2, FUN=PI) %>% t
    bk_tmp <- tibble(
      p = mean_p,
      p5 = p89interval[,1],
      p94 = p89interval[,2]
    ) %>% 
      mutate(D = row_number(), 
             xpos = str_pad((D-1) * nkids + k, 3, pad = 0, ), 
             K = as.character(k))
    return(bk_tmp)
  }
)


# Plot
bk_effect_plot <- ggplot(bk_to_plot) +
  geom_segment(
    aes(x=D, xend=D, y=p5, yend=p94, color=K), 
    alpha = 0.4, linewidth=2, lineend = "round"
  ) +
  geom_point(aes(x=D, y=p, color=K)) +
  geom_hline(aes(yintercept=mean(one_kid_samples)), linetype=2) +
  scale_y_continuous(limits = c(0, 100)) +
  facet_grid(K ~.) +
  labs(x="District", y="Percentual points (pp)", 
       title = "Direct effect of Surviving kids on contraceptive per district") +
  theme_minimal()
bk_effect_plot

```

```{r}
#| fig-height: 15
bk_effect_plot <- ggplot(bk_to_plot) +
  geom_segment(
    aes(x=xpos, xend=xpos, y=p5, yend=p94, color=K),
    alpha = 0.4, linewidth=2, lineend = "round"
  ) +
  geom_point(aes(x=xpos, y=p, color=K)) +
  geom_hline(aes(yintercept=mean(one_kid_samples)), linetype=2) +
  scale_y_continuous(limits = c(0, 100)) +
  facet_wrap(D ~., scales = "free_x") +
  labs(x="District", y="Percentual points (pp)", 
       title = "Direct effect of Surviving kids on contraceptive per district") +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(), 
        axis.text.x = element_blank())
bk_effect_plot
```

```{r}
kids <- c(1:4)
nkids <- length(kids)
contrast_to_plot <- map_df(
  kids, 
  .f=function(k){
    mbK <- m2_posterior$bK %>% rep(61) %>% matrix(ncol=61)
    bK_samples <- with(
      m2_posterior,
      if(k == 1){
        inv_logit(a + bU) * 100
      }else{
        inv_logit(a + bU + mbK * apply(cbind(0,delta)[,1:k],1, sum)) * 100
      }
    )
    bK_cluster_samples <- with(
      m3_posterior,
      if(k == 1){
        inv_logit(a + bU) * 100
      }else{
        inv_logit(a + bU + bK * apply(cbind(0,delta)[,1:k],1, sum)) * 100
      }
    )
    # Build data frame
    contrast <- bK_cluster_samples - bK_samples
    mean_p <- contrast  %>% apply(MARGIN=2, FUN=mean)
    p89interval <- contrast %>% apply(MARGIN=2, FUN=PI) %>% t
    bk_tmp <- tibble(
      p = mean_p,
      p5 = p89interval[,1],
      p94 = p89interval[,2]
    ) %>% 
      mutate(D = row_number(), 
             xpos = str_pad((D-1) * nkids + k, 3, pad = 0, ), 
             K = as.character(k))
    return(bk_tmp)
  }
)


# Plot
bk_effect_plot <- ggplot(contrast_to_plot) +
  geom_segment(
    aes(x=D, xend=D, y=p5, yend=p94, color=K), 
    alpha = 0.4, linewidth=2, lineend = "round"
  ) +
  geom_point(aes(x=D, y=p, color=K)) +
  geom_hline(aes(yintercept = 0), linetype=2) +
  facet_grid(K ~.) +
  labs(x="District", y="Percentual points (pp)", 
       title = "Constrast on the effect of living children when causal effect of children variesby district") +
  theme_minimal()
bk_effect_plot
```

```{r}
#| fig-height: 15
ggplot(contrast_to_plot) +
  geom_segment(
    aes(x=xpos, xend=xpos, y=p5, yend=p94, color=K),
    alpha = 0.4, linewidth=2, lineend = "round"
  ) +
  geom_point(aes(x=xpos, y=p, color=K)) +
  geom_hline(aes(yintercept=0), linetype=2) +
  facet_wrap(D ~., scales = "free_x") +
  labs(x="District", y="Percentual points (pp)", 
       title = "Constrast on the effect of living children when causal effect of children varies by district") +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(), 
        axis.text.x = element_blank())

```