---
title: "Homework - Week 01"
format: html
editor: visual
---

```{r}
library(magrittr)
library(glue)
library(ggplot2)
```


# WEEK 1 PROBLEMS

## Problem 00

For your own good, it would be helpful to review the EASY problems at the end of Chapters 1, 2 and 3 (2nd edition). The answers
are in the solutions guide.

## Problem 01

1. Suppose the globe tossing data (Lecture 2, Chapter 2) had turned out to be 4 water and 11 land. Construct the posterior distribution.

```{r}
compute_posterior <- function(W, L, precision_step=100){
  N <- W + L
  p <- seq(0, 1, length.out = 100)
  
  # P(p|W,L) = P(W,L|p)P(p) / P(W,L)
  likelihood <- sapply(p, function(x) dbinom(W, N, x))
  posterior = likelihood * p / sum(likelihood * p)
  return(posterior)
}
```


```{r 01_solution}
W <- 4
L <- 11
posterior = compute_posterior(W, L, 100)

# Plot function approximation
ggplot() +
  geom_line(aes(x=p, y=posterior)) +
  geom_point(aes(x=p, y=posterior)) +
  labs(x = "Proportion of water",
       y = "Posterior probability",
       title = glue("Posterior probability for W = {W} and L = {L}"))

```

```{r 01_book_example}
W <- 6
L <- 3
posterior = compute_posterior(W, L, 100)

ggplot() +
  geom_line(aes(x=p, y=posterior)) +
  geom_point(aes(x=p, y=posterior)) +
  labs(x = "Proportion of water",
       y = "Posterior probability",
       title = glue("Posterior probability for W = {W} and L = {L}"))

```
## Problem 02

2. Using the posterior distribution from 1, compute the posterior predictive distribution for the next 5 tosses of the same globe. I recommend you use the sampling method.

## Problem 03

3. Use the posterior predictive distribution from 2 to calculate the probability of 3 or more water samples in the next 5 tosses.

## Problem 04
4-OPTIONAL. This problem is an optional challenge for people who are taking the course for a second or third time. Suppose you observe W = 5 water points, but you forgot to write down how many times the globe was tossed, so you don’t know the number of land points L. Assume that p = 0.7 and compute the posterior distribution of the number of tosses N. Hint: Use the binomial distribution.

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
