---
title: "Homework - Week 01"
format: html
editor: visual
---

```{r}
library(dplyr)
library(glue)
library(ggplot2)
options(scipen = 999)
```


# WEEK 1 PROBLEMS

## Problem 00

For your own good, it would be helpful to review the EASY problems at the end of Chapters 1, 2 and 3 (2nd edition). The answers
are in the solutions guide.

## Problem 01

1. Suppose the globe tossing data (Lecture 2, Chapter 2) had turned out to be 4 water and 11 land. Construct the posterior distribution.

```{r}
compute_posterior <- function(W, L, precision_step=100){
  N <- W + L
  p <- seq(0, 1, by = 1/precision_step)
  
  # P(p|W,L) = P(W,L|p)P(p) / P(W,L)
  likelihood <- sapply(p, function(x) dbinom(W, N, x))
  posterior = likelihood / sum(likelihood)
  
  return(tibble(p, likelihood, posterior))
}
```


```{r 01_book_example}
W <- 6
L <- 3
precision <- 100
df_posterior = compute_posterior(W, L, precision)

ggplot(data=df_posterior) +
  geom_line(aes(x=p, y=posterior)) +
  geom_point(aes(x=p, y=posterior)) +
  labs(x = "Proportion of water",
       y = "Posterior probability",
       title = glue("Posterior probability for W = {W} and L = {L}"))

```

```{r 01_solution}
W <- 4
L <- 11
precision <- 10
df_posterior = compute_posterior(W, L, precision)

df_posterior %>% 
  mutate_all(.f=round, digits=3)
```

```{r 01_solution}
W <- 4
L <- 11
precision <- 10
df_posterior = compute_posterior(W, L, precision)

df_posterior %>% 
  mutate_all(.f=round, digits=3)
```

```{r}
# Plot function approximation
ggplot(data=df_posterior) +
  geom_line(aes(x=p, y=posterior)) +
  geom_point(aes(x=p, y=posterior)) +
  labs(x = "Proportion of water",
       y = "Posterior probability",
       title = glue("Posterior probability for W = {W} and L = {L}"))

```


## Problem 02

2. Using the posterior distribution from 1, compute the posterior predictive distribution for the next 5 tosses of the same globe. I recommend you use the sampling method.

```{r 02_solution}
W <- 4
L <- 11
precision <- 100
df_posterior = compute_posterior(W, L, precision)

next_sample_size = 5
nsims = 1e4

# Estimate probability
p <- df_posterior$p
posterior <- df_posterior$posterior
sampled_p = sample(p, size=nsims, replace=T, prob = posterior)

pred_tosses = sapply(
  sampled_p,
  function(q){                   
    sum(
      sample(
        c("W", "L"), 
        size = next_sample_size,
        replace = T, 
        prob = c(q, 1-q)
      )=="W"
    )
  }
)

sample_counts <- table(pred_tosses) %>% as_tibble()
sample_counts

ggplot(data = sample_counts) +
  geom_point(aes(x=pred_tosses, y=n)) +
  geom_segment(aes(x=pred_tosses, xend=pred_tosses,
                   y=0, yend=n),
               size=1.5
  ) +
  labs(x = "Number of \"W\"",
       y = "Sample counts",
       title = glue("Posterior predictive distribution for the next 5 tosses (W = {W} and L = {L})"))
```

## Problem 03

3. Use the posterior predictive distribution from 2 to calculate the probability of 3 or more water samples in the next 5 tosses.

```{r}
#P(W>=3)
sample_counts %>% 
  mutate(prob= n / sum(n)) %>% 
  filter( as.numeric(pred_tosses) >= 3) %>% 
  summarise(prob_gte_3= sum(prob))
```


## Problem 04
4-OPTIONAL. This problem is an optional challenge for people who are taking the course for a second or third time. Suppose you observe W = 5 water points, but you forgot to write down how many times the globe was tossed, so you don’t know the number of land points L. Assume that p = 0.7 and compute the posterior distribution of the number of tosses N. Hint: Use the binomial distribution.

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
